
This thesis would not have been possible without the support of many people, a few of whom I have space to thank below.

\subsection*{Individual}

To my supervisors, Simon Myers and Jonathan Marchini, for taking me on as a student, for providing guidance and feedback, for having the patience to afford me the independence of carving my own path, and for your infectious enthusiasm for science.

To my collaborators and colleagues, Min Jung and Don Conrad; Emmanuelle Bitoun; Anjali Hinch, Gang Zhang, Peter Donnelly and Ben Davies; for generously sharing data and ideas.

To my first rotation supervisor Benjamin Schuster-Böckler for guiding me in my initial foray into genomics.

To my advisors and examiners, Ian Henderson, Gerton Lunter, Luke Jostins, Chris Holmes, and John Todd for their valuable advice and feedback on my work.

To Julian Knight and Isabel Schmidt, for supporting the Genomic Medicine and Statistics programme and community as well as their support and advice to me personally.

To my coursemates and officemates, for their comradeship and support.

To my parents, for their immeasurable love and support.

\subsection*{Institutional}

To the Wellcome Trust for providing generous funding through the Genomic Medicine and Statistics programme (grant 109109/Z/15/Z).

To the Wellcome Center of Human Genetics and the Department of Statistics for providing office space and talks, and a supportive research environment.

To Linacre College, for providing accommodation, dining, and fitness facilities.
